Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-16 ms/spaceness.tex
full discussion draft

\documentclass[10pt,twoside,lineno]{gsajnl}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\setstretch{1}
% Use the documentclass option 'lineno' to view line numbers

\articletype{preprint} % article type
% {inv} Investigation 
% {gs} Genomic Selection
% {goi} Genetics of Immunity 
% {gos} Genetics of Sex 
% {mp} Multiparental Populations

% adding commenting commands
\newif\ifcomments
\commentstrue
%\commentsfalse
\newcommand{\cjb}[1]{\ifcomments{{\color{blue} \it (#1)}}\else{}\fi}
\newcommand{\plr}[1]{\ifcomments{{\color{purple} \it (#1)}}\else{}\fi}
\newcommand{\ak}[1]{\ifcomments{{\color{red} \it (#1)}}\else{}\fi}

@@

\author[$\ast$,1]{C.J. Battey}
\author[$\ast$]{Peter Ralph}
\author[$\ast$]{Andrew Kern}

@@

\runningtitle{SPACENESS} % For use in the footer

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many models used to infer selection, phenotypic associations, or past population sizes assume that samples are a random draw from a panmictic population. In reality most populations are structured to some degree across space, but the extent to which this model violation biases common genomic analyses is not well known. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies. We find that estimates of past population sizes and many common population genetic summary statistics vary significantly with both dispersal and sampling, with most bias occurring when neighborhood sizes are below 100 and sampling is spatially clustered. Variation in the site frequency spectrum and Tajima's D is greater across sampling strategies than over the range of demographic parameters we simulated, pointing to the need to carefully account for sampling when interpreting results. Last we show that the combination of isolation by distance and spatially correlated environments causes genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.  
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models to describing are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a mixture of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will not represent a random sample of the genetic variation present in the whole population. This is as true at the local scale as it is between the more differentiated units typically analyzed in phylogeographic studies. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

\ak{we should start by saying that we don't understand how simple summaries will respond to continuous space model}
This issue is particularly important for analyses of polygenic selection, because the allele frequency variation caused by population structure is similar to the pattern expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was badly confounded by population structure \citep{Sohail2018,Berg2018}. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.

Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are likely to fail.

@@

\ak{i moved the following section into M\&M. makes more sense here i think}

\subsection{Modeling Evolution in Continuous Space}
The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

Though some aspects of continuous populations are fairly well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible. Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).

@@

\subsection{Genealogical Parameters}
We recorded the census population size and estimated the variance in number of offspring and generation time \ak{need to explain why this is needed} for all simulations. To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean. To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.

@@

\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $pi$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were meant to roughly reflect the distribution of height across Europe \ak{how so? need more details here. Was it matched to mean and variance?}, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10. Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map. Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map. Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map. \ak{in addition to the verbal treatment above I would list the formulae that you are using for the phenotypes here}

@@

\subsection{Genealogical Parameters and Run Times}
Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number. In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantitives like the generation time depend on population parameters such as the dispersal distance and population density over the landscape. We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size which itself is a function of $\sigma$ (Figure \ref{fig:genparams}). There are a few things to note. First all three quantities are non-linear with respect to neighborhood size. Census size largely declines for both the spatial and random mating models, however for spatial models this decline only begins for neighborhood size $\geq 10$. By a neighborhood size $\geq 100$ the spatial and random mating models are largely indistinguishable from one another, a sign that our simulations are performing as expected. Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.

Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation. To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is itself determined by the parameterization of the model. Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$. Under the range of neighborhood sizes that we examined here, generation times between the random mating and spatial models are never quite equivalent-- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

Last we looked at the effect of neighborhood size on the variance in number of offspring in the population-- a key parameter determining the effective population size. Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.

@@

\subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}
One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models. Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation as strongly, if not more so, than the underlying population dynamics.

@@

\subsubsection{Effects of Sampling on Random Mating and Spatial Models}
To summarize the effect of sampling strategy on genetic diversity in random mating and spatial models we also asked if summary statistics varied significantly across sampling regimes. In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models. Essentially, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.

@@

In Figure \ref{fig:demography} we show inferred population size histories from our simulations as a function of neighborhood size binned in to one of four categories. In general we observe that demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). This is consistent with our observations of the SFS from which Stairwayplot is doing its inference. Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates. More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies. Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$. While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.

To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation. Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process. Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}). In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot. This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.

@@

When PCA positions are included as covariates for GWAS the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C). At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments. Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}. This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B). Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.

>>>>

\documentclass[11pt,twoside,lineno]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\setstretch{1.5}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

% adding commenting commands
\newif\ifcomments
\commentstrue
%\commentsfalse
\newcommand{\cjb}[1]{\ifcomments{{\color{orange} \it (#1)}}\else{}\fi}
\newcommand{\plr}[1]{\ifcomments{{\color{purple} \it (#1)}}\else{}\fi}
\newcommand{\ak}[1]{\ifcomments{{\color{red} \it (#1)}}\else{}\fi}

@@

\author[$\ast$,1]{C.J. Battey}
\author[$\ast$]{Peter L. Ralph}
\author[$\ast$]{Andrew D. Kern}

@@

\runningtitle{Effects of Continuous Space} % For use in the footer

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many methods of analyzing genetic data assume that samples are a random draw from a random-mating population, but are applied to real populations that are structured over space by limited dispersal. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies. We find that estimates of population size trajectories and many common summary statistics vary significantly with dispersal, with most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered. In association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome, and this bias is only partially corrected by regressing out principal components positions in the analysis. Last we describe the genealogical process leading to spatial effects we observed on simulated gentoypes, summarize estimates of neighborhood size from empirical studies across a range of organisms, and discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a set of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

Our understanding of how many common summary statistics respond to variation in dispersal distance is limited. Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}). However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals. Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these genealogical summaries may also reflect variation in the scale of spatial interactions both over time and across populations. Similarly, the distribution of haplotype block lengths (i.e. stretches of identical sequence) among individuals is sometimes interpreted as reflecting historic demography \citep{Harris2013} or evidence of selection \citep{Garud2015}. However as this distribution is also an indirect estimate of the branch lengths of the sample genealogy it is also sensitive to variation in dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.

A related issue is the spatial distribution of sampling effort. Though individuals in a random-mating population are exchangeable, those in a continuously structured population are not. If sampling is clustered at a series of points on the landscape, as is common in empirical studies because of the practical difficulties of sample collection, samples drawn from a spatially structured population may partially reflect spatial biases in the sampling process rather than in the underlying population. In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, but the effects of this process on many downstream inference procedures is unknown. \cjb{must be more out there on sampling -- any recs for papers to read+cite?}

The issue of fine-scale structure for putatively "within-population" analyses is particularly important for studies of polygenic selection. This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to bias in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.

Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.

@@

\subsection{Modeling Evolution in Continuous Space}
The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

Though some aspects of continuous populations are well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible. Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).

@@

\subsection{Genealogical Parameters}
Because basic elements of our simulations like census population size and generation time are not directly parameterized in our model (instead they emerge from the interaction of parameters including average population density, average lifespan, and the spatial scale of interaction), we first monitored a range of genealogical parameters in all simulations. We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations. To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean. To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
\label{fig:samplemap}
\end{figure}

\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10.
\begin{equation}
    P \sim \mathcal{N}(110,10)
\end{equation}
Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map. If $x$ is an individual's x coordinate on a 50x50 landscape, the phenotype is then:
\begin{equation}
    P \sim \mathcal{N}(110+\frac{2x}{5},10)
\end{equation}
Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map. 
\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(130,10),& \text{if } x< 20 \text{ and } y<20\\
    \mathcal{N}(110,10),& \text{otherwise}\\
\end{cases}
\end{equation}

Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map.

\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(130,10),& \text{if } d < 3\\
    \mathcal{N}(110,10),& \text{otherwise}\\
\end{cases}
\end{equation}
where $d$ is the distance to the nearest patch.

@@

\subsection{Genealogical Parameters}
Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number. In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantities like the generation time depend on life history parameters such as the dispersal distance and carrying capacity of the landscape. We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size, which itself is a function of $\sigma$ (Figure \ref{fig:genparams}). There are a few things to note. First all three quantities are non-linear with respect to neighborhood size. Census size largely declines as neighborhood size increases for both the spatial and random mating models. However for spatial models this decline only begins for neighborhood size $\geq 10$. By a neighborhood size $\geq 100$ the spatial and random mating models are indistinguishable from one another, a sign that our simulations are performing as expected. Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.

\begin{figure}[htbp]
\centering
\includegraphics{figures/pop_params.pdf}
\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
\label{fig:genparams}
\end{figure}

Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation. To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is only indirectly determined by the parameterization of density-dependent competition the model. Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$. Interestingly under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

Last we looked at the effect of neighborhood size on the variance in number of offspring in the population -- a key parameter determining the effective population size. Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.

@@

\subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}
One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models. Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation at least as strongly as the underlying population dynamics.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
\caption{Site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

\subsubsection{Effects of Sampling on Random Mating and Spatial Models}
To summarize the effect of sampling strategy on genetic diversity in random mating and spatial models we also asked if summary statistics varied significantly across sampling regimes. In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models. In summary, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.

@@

In Figure \ref{fig:demography} we show inferred population size histories from our simulations as a function of neighborhood size binned in to one of four categories. In general we observe that demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). This is consistent with our observations of the SFS from which Stairwayplot is doing its inference. Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates. More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies. Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$. While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.  
To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation. Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process. Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}). In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot. This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range. The thick line is a rolling mean and thin lines are individual model fits. Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
\label{fig:demography}
\end{figure}

@@

When PCA positions are included as covariates to control for population structure in GWAS the vast majority of SNPs no longer surpass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C). At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments. Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}. This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B). Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.
